Kindly Bent to Free Us

### 2019-11-15 kindly/lang.tex
Make the goal of the examples clearer.

In the examples presented in \cref{motivation}, region annotations are optional.
In our formalization, regions must be syntactically explicit and annotated with
a nesting index and a scoped variable.
We now define a best-effort
syntactic code transformation $\RannotT{p}{p'}$ which
annotates programs automatically.
The input $p$ is a program with optional region annotations of the form $\region[]{}{e}$.
The output $p'$ is a program with explicit annotations
of the form $\region{\Sone x\BORROW}{e}$.
We give an informal presentation of our code transformation. The complete
definition is given in \cref{appendix:regionannot}. This code transformation
aims to find, for each borrow $\borrow{x}$, the biggest region satisfying
the following rules:
\begin{enumerate}
\item The region should contain at least $\borrow{x}$.
\item The region should never be larger than the scope of $x$.
\item An exclusive borrow $\borrow[\kaff]{x}$ should never share a region with any other borrow of $x$.
\item A use of $x$ can never be in the region of $\borrow{x}$.
\end{enumerate}
Following these rules, we start from the leafs and grow the region
associated to each borrow until it can't grow anymore, either due to a binding
or to conflicting uses.

@@

Since the only use of $a$ is a borrow,
the region covers the entire lexical scope of $a$. For $x$ however, there are
multiple conflicting borrow and we must be more careful. We place a first
region around the exclusive borrow and its function application,
and a second region around both shared borrows. This placement of region
is optimal, as making any of the region bigger would cause a borrowing
error.

Naturally, these rules are best effort and it's sometimes necessary for the
programmer to annotate regions explicitly. We take such manual annotations
into account as an upper bound on all the contained regions.
In the following program, a manual annotation has been inserted to ensure
no borrow enters the reference $r$:

@@

In order to deconstruct pairs in the presence of linear types, we require
more than projection operators. Indeed, given a linear pair $x$,
$\operatorname{fst} x$ would consume $x$ without giving access to the second
member.
Furthermore, we would like to be able to deconstruct a borrow of a pair
and obtain borrows of the components.
For these purposes, we provide the pattern matching construct
$\matchin{x,x'}{e_1}{e_2}$. The typing rule is shown in \cref{selectrules:matching}.
A syntactic marker $\etransfm$ indicates if it applies to
a pair ($\etransfm = \operatorname{id}$) or a borrow ($\etransfm = \&^\BORROW$).
If $\etransfm = \operatorname{id}$, it operators a regular destruct on pair.
Otherwise, $e_1$ is expected to be of type
$\borrowty{k}{\tyPair{\tau_1}{\tau'_1}}$
while $x$ and $x'$ are of type respectively
$\borrowty{k}{\tau_1}$ and $\borrowty{k}{\tau'_1}$.

>>>>

In the examples presented in \cref{motivation}, region annotations are optional.
In our formalization, regions must be syntactically explicit and annotated with
a nesting index and a scoped variable.
We now define a best-effort
syntactic code transformation $\RannotT{p}{p'}$ which
annotates programs automatically.
The input $p$ is a program with optional region annotations of the form $\region[]{}{e}$.
The output $p'$ is a program with explicit annotations
of the form $\region{\Sone x\BORROW}{e}$.
We give an informal presentation of our code transformation. The complete
definition is given in \cref{appendix:regionannot}. This code transformation
aims to find, for each borrow $\borrow{x}$, the biggest region satisfying
the following rules:
\begin{enumerate}
\item The region should contain at least $\borrow{x}$.
\item The region should never be larger than the scope of $x$.
\item An exclusive borrow $\borrow[\kaff]{x}$ should never share a region with any other borrow of $x$.
\item A use of $x$ can never be in the region of $\borrow{x}$.
\end{enumerate}
Following these rules, we start from the leafs and grow the region
associated to each borrow until it can't grow anymore either due to a binding
or to conflicting uses.

@@

Since the only use of $a$ is a borrow,
the region covers the entire lexical scope of $a$. For $x$ however, there are
multiple conflicting borrow and we must be more careful. We place a first
region around the exclusive borrow and its function application,
and a second region around both shared borrows. This placement of region
is optimal, as making any of the region bigger would cause a borrowing
error.
Notes that indices of regions are well nested, as the inner region have
higher indices that the outer ones.

Naturally, these rules are best effort and programmers must sometime
annotate regions explicitly. We take such manual annotations
into account as an upper bound on the regions of all contained borrows.
In the following program, a manual annotation has been inserted to ensure
no borrow enters the reference $r$:

@@

In order to deconstruct pairs in the presence of linear types, we require
more than projection operators. Indeed, given a linear pair $x$,
$\app{\operatorname{fst}}{x}$ would consume $x$ without giving access to the second
member.
Furthermore, we would like to be able to deconstruct a borrow of a pair
and obtain borrows of the components.
For these purposes, we provide the construct
$(\matchin{x,x'}{e_1}{e_2})$ whose
typing rule is shown in \cref{selectrules:matching}.
A syntactic marker $\etransfm$ indicates if it applies to
a pair ($\etransfm = \operatorname{id}$) or a borrow ($\etransfm = \&^\BORROW$).
If $\etransfm = \operatorname{id}$, the typing simplifies to
a regular destruct on pairs.
Otherwise, $e_1$ is expected to be of type
$\borrowty{k}{\tyPair{\tau_1}{\tau'_1}}$
while $x$ and $x'$ are of type respectively
$\borrowty{k}{\tau_1}$ and $\borrowty{k}{\tau'_1}$.

### 2019-11-15 kindly/motivation.tex
Make the goal of the examples clearer.

\lang{} is amenable to a functional resource-passing style as it is
common in functional encodings of session types (e.g.,
\cite{DBLP:journals/jfp/Padovani17}; see also
\cref{sec:session-linearity} in the supplement). But it really shines 
when manipulating mutable resources like buffers or connection pools
using an imperative programming style. 
%
To support this usage pattern of linearity, we rely on
the notion of borrowing \cite{DBLP:conf/popl/BoylandR05}.

\subsection{Imperative programming}
\label{sec:imper-progr}

A first example is an API for mutable linear arrays
(\cref{sig:array}).
The intention of this API is to safely handle manual allocation and
deallocation of arrays that may contain affine elements.
One would first use \lstinline/create n v/ to create
an (unrestricted) array of size \lstinline/n/ initialized with value
\lstinline/v/. This value must be unrestricted as it is duplicated to
initialize all array elements. Using the \lstinline/map/ function we
can transform the unrestricted elements into linear (affine)
ones. This array can be processed further using \lstinline/map/ (or \lstinline/set/ if
affine). The \lstinline/get/ function is only applicable if the element type
is unrestricted as one element is duplicated. The \lstinline/length/
function is always applicable. To \lstinline/free/ an array the elements must be
affine.

@@

\begin{figure}[tp]
  \centering
  \begin{subfigure}[t]{1\linewidth}
\begin{lstlisting}
module Array : sig
  type ('a : 'k) t : lin
  val create : ('a : un) => int -> 'a -> 'a t
  val free : ('a : aff) => 'a t -> unit
  val length : &('a t) -> int
  val get : ('a : un) => &('a t) * int -> 'a
  val set : ('a : aff) => &!('a t) * int * 'a -> unit
  val map : (&'a -> 'b) -> &('a t) -> 'b t
end
\end{lstlisting}
    \vspace{-15pt}
    \caption{Affine arrays}
    \label{sig:array}
  \end{subfigure}

  \begin{subfigure}[t]{1\linewidth}
\begin{lstlisting}
let mk_fib_array n =
  let a = create n 1 in
  for i = 2 to n - 1 do
    let x = get (&a, i-1) + get (&a, i-2) in(*@\label{ex:array:get}*)
    set (&!a, i, x)(*@\label{ex:array:set}*)
  done;
  a(*@\label{ex:array:return}*)
# mk_fib_array : int -> int Array.t
\end{lstlisting}
    \vspace{-10pt}
    \caption{Example of use of linear arrays}
    \label{ex:array}
\begin{lstlisting}[firstnumber=2]
  let a = create n 1 in
  for i = 2 to n - 1 do {|
    let x = {| get &a (i-1) + get &a (i-2) |} in(*@\label{ex:array:region:get}*)
    set &!a i x
  |} done;
\end{lstlisting}
    \vspace{-10pt}
    \caption{Excerpt of \cref{ex:array} with explicit regions}
    \label{ex:array:region}
  \end{subfigure}
\end{figure}

@@

\subsection{Hybrid programming style}

@@

\cref{sig:hybarray} contains an API \lstinline/HybArray/
that enables using mutable and immutable modifications to
the board through affine types and borrows.
The signature differs slightly from the \lstinline/Array/
signature. For simplicity, the array elements are unrestricted, but
the structure itself remains linear so as to be implemented in terms of \lstinline/Array/.
The in-place mutation function \lstinline/set_mut/ with type
\lstinline/&!('a t) * int * 'a -> unit/ works on a exclusive borrow  whereas the persistent
\lstinline/set/ operation has type \lstinline/&('a t) * int * 'a -> 'a t/. It
takes an shared borrow because it only reads from the
argument array, but returns a fresh,  modified structure.
The code in \cref{ex:cow} contains a very simple implementation of
\lstinline/HybArray/ that represents hybrid arrays
as regular arrays and uses copy-on-write for persistent
modifications. The function
\lstinline/mapi : (int -> &'a -> 'b) -> &('a t) -> 'b t /
is a simple variation on \lstinline/Array.map/ where the mapping function
takes the position of the element.
% A more elaborate implementation could use persistent arrays
% \cite{DBLP:conf/ml/ConchonF07}.

@@

create : (unit -> 'a) -> 'a pool
consume : ('a -> unit) -> 'a pool -> unit
use : ('a_1:'k_1),('a_2:'k_2),('k_2 <= aff_r) =>
  &('a_1 pool) -> (&!(aff_r1,'a_1) -{lin}> 'a_2) -{'k_1}> 'a_2
\end{lstlisting}
    \vspace{-10pt}
    \caption{Interface}
    \label{intf:pool}

>>>>

\lang{} is amenable to a functional resource-passing style as it is
common in functional encodings of session types (e.g.,
\cite{DBLP:journals/jfp/Padovani17}; see also
\cref{sec:session-linearity} in the supplement). But it really shines 
when manipulating mutable resources like buffers or connection pools
using a mixed functional and imperative programming style.
%
To support this usage pattern of linearity, we rely on
the notion of borrowing \cite{DBLP:conf/popl/BoylandR05} which
we demonstrate through a series of examples:
\begin{itemize}
\item An API for linear arrays to demonstrate simple borrowing and imperative programming.
\item
  A backtracking Sudoku solver to demonstrate
  \lang's ability to build safe hybrid functional/imperative data-structures.
\item A pool of linear objects to demonstrate advanced usage of regions.
\end{itemize}

\subsection{Imperative programming with linear arrays}
\label{sec:imper-progr}

We now showcase an API for mutable linear arrays
(\cref{sig:array}).
The intention of this API is to safely handle manual allocation and
deallocation of arrays that may contain affine elements.
One would first use \lstinline/create n v/ to create
an (unrestricted) array of size \lstinline/n/ initialized with value
\lstinline/v/. This value must be unrestricted as it is duplicated to
initialize all array elements. Using the \lstinline/map/ function we
can transform the unrestricted elements into linear (affine)
ones. This array can be processed further using \lstinline/map/ (or \lstinline/set/ if
affine). The \lstinline/get/ function is only applicable if the element type
is unrestricted as one element is duplicated. The \lstinline/length/
function is always applicable. To \lstinline/free/ an array the elements must be
affine.

@@

\begin{figure}[tp]
  \centering
  \begin{subfigure}[t]{1\linewidth}
\begin{lstlisting}
module Array : sig
  type ('a : 'k) t : lin
  val create : ('a : un) => int -> 'a -> 'a t
  val free : ('a : aff) => 'a t -> unit
  val length : &('a t) -> int
  val get : ('a : un) => &('a t) * int -> 'a
  val set : ('a : aff) => &!('a t) * int * 'a -> unit
  val map : (&'a -> 'b) -> &('a t) -> 'b t
end
\end{lstlisting}
    \vspace{-15pt}
    \caption{Signature}
    \label{sig:array}
  \end{subfigure}

  \begin{subfigure}[t]{1\linewidth}
\begin{lstlisting}
let mk_fib_array n =
  let a = create n 1 in
  for i = 2 to n - 1 do
    let x = get (&a, i-1) + get (&a, i-2) in(*@\label{ex:array:get}*)
    set (&!a, i, x)(*@\label{ex:array:set}*)
  done;
  a(*@\label{ex:array:return}*)
# mk_fib_array : int -> int Array.t
\end{lstlisting}
    \vspace{-10pt}
    \caption{Example of use}
    \label{ex:array}
\begin{lstlisting}[firstnumber=2]
  let a = create n 1 in
  for i = 2 to n - 1 do {|
    let x = {| get &a (i-1) + get &a (i-2) |} in(*@\label{ex:array:region:get}*)
    set &!a i x
  |} done;
\end{lstlisting}
    \vspace{-10pt}
    \caption{Excerpt of \cref{ex:array} with explicit regions}
    \label{ex:array:region}
  \end{subfigure}
  \vspace{-5pt}
  \caption{Linear arrays}
  \label{ex:array}
\end{figure}

@@

\subsection{Solving sudokus with hybrid data-structures}

@@

\cref{sig:hybarray} contains an API \lstinline/HybArray/
that enables using mutable and immutable modifications to
the board through affine types and borrows.
The signature differs slightly from the \lstinline/Array/
signature. For simplicity, the array elements are unrestricted, but
the structure itself remains linear so as to be implemented in terms of \lstinline/Array/.
The in-place mutation function \lstinline/set_mut/ with type
\lstinline/&!('a t) * int * 'a -> unit/ works on a exclusive borrow  whereas the persistent
\lstinline/set/ operation has type \lstinline/&('a t) * int * 'a -> 'a t/. It
takes a shared borrow because it only reads from the
argument array, but returns a fresh,  modified structure.
The code in \cref{ex:cow} contains a very simple implementation of
\lstinline/HybArray/ that represents hybrid arrays
as regular arrays and uses copy-on-write for persistent
modifications. The function
\lstinline/mapi : (int -> &'a -> 'b) -> &('a t) -> 'b t /
is a simple variation on \lstinline/Array.map/ where the mapping function
takes the position of the element.
% A more elaborate implementation could use persistent arrays
% \cite{DBLP:conf/ml/ConchonF07}.

@@

create : (unit -> 'a) -> 'a pool
consume : ('a -> unit) -> 'a pool -> unit
use : ('a_1:'k_1),('a_2:'k_2),('k_2 <= aff_r) =>
  &('a_1 pool) -> (&!(aff_r1,'a_1) -{lin}> 'a_2) -{'k_1}> 'a_2
\end{lstlisting}
    \vspace{-10pt}
    \caption{Signature}
    \label{intf:pool}
